Archaic introgression and the distribution of shared variation under stabilizing selection

### 2024-08-19 doc/manuscript.tex
paper edits

\usepackage[letterpaper,margin=1in]{geometry}
\usepackage{lineno}   %% <- no mathlines option
\usepackage{amsmath}  %% <- after lineno
\usepackage{etoolbox} %% <- for \cspreto, \csappto
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{color}
\usepackage{url}

@@

\DeclareMathOperator{\Var}{Var}
\newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
\newcommand{\E}{\mathbb{E}}
\usepackage{xspace}
\newcommand{\moments}{\texttt{moments}\xspace}
\newcommand{\fwdpy}{\texttt{fwdpy11}\xspace}
\newcommand{\tskit}{\texttt{tskit}\xspace}

@@

\title{Archaic introgression and the distribution
of shared functional variation under stabilizing selection}
\author{APR}
\date{\today}

@@

    Many phenotypic traits are under stabilizing selection, which maintains
    individual phenotypic values near some optimum. Classical theory shows that
    at loci contributing to the genetic variance of a trait under stabilizing
    selection, the minor allele is selected against via symmetric
    underdominance. Most studies of the dynamics of the genetic architecture
    and variance of a trait under stabilizing selection have focused on single
    populations. However, natural populations are typically structured in some
    way, and admixture and introgression are common occurances, in particular
    in human evolution. Because stabilizing selection results in selection
    against the minor allele at a trait-affecting locus, the alleles from the
    minor parental ancestry will be selected against after admixture. Here, we
    show that the site frequency spectrum can be used to model the genetic
    architecture of such traits, allowing for complex multi-population
    histories including admixture. We develop a simple model with linkage to a
    neutral locus to show that introgressed ancestry is expected to be depleted
    around trait-affecting alleles. Using this model and individual-based
    simulations, we show that introgressed ancestry deserts tend to form around
    such loci. When introgression between two diverged populations occurs in
    both directions, as has been inferred between humans and Neanderthals,
    introgressed ancestry deserts will often be shared across populations.
    Stabilizing selection for a shared phenotypic optimum may therefore explain
    recent observations in which regions of depleted human-introgressed
    ancestry in the Neanderthal genome overlap with Neanderthal-ancestry
    deserts in humans.

@@

Genomic surveys of natural systems show that historical admixture among
diverged populations or closely related taxa commonly occurs
\citep{brandvain2014speciation, skoglund2015ancient, suvorov2022widespread} and
is potentially widespread in primate \citep{tung2017contribution,
sorensen2023genome} and hominin \citep{wolf2018outstanding, peter2020100}
evolution. Admixture may therefore be an important driver of phenotypic and
molecular variation and shapes the genetic basis of complex traits. In humans,
archaic introgression involving Neanderthals and Denisovans has attracted
considerable attention, including efforts to describe the historical processes
leading to observed distributions of introgressed DNA in present-day
populations \citep{prufer2014complete, villanea2019multiple,
chen2020identifying} and the contribution of introgressed variation to
quantitative traits \citep{sankararaman2016combined, wei2023lingering}.

Once introduced through admixture, introgressed alleles may be selected for or
against. Some introgressed haplotypes were likely positively selected in modern
humans \citep{huerta2014altitude, racimo2017signatures, enard2018evidence,
gower2021detecting}, possibly due to locally adaptive variation that provided
fitness advantages as humans encountered novel environments. Despite some cases
of adaptive introgression, mostly introduced alleles were likely to have been
selected against in humans \citep{harris2016genetic, juric2016strength}. Since
Neanderthal and Denisovan population sizes were relatively small for hundreds
of thousands of years, theory predicts they would have accumulated fixed
deleterious variation at an increased rate. Introgressed haplotypes loaded with
more deleterious mutations would then have quickly been selected against after
admixture. Mapping the distribution of Neanderthal-introgressed haplotypes in
humans shows a reduction of Neanderthal-related ancestry in coding and
regulatory regions \citep{petr2019limits, telis2020selection,
yermakovich2023long}. These ``deserts'' of Neanderthal ancestry support the
hypothesis that introgressed functional alleles were selected against
\citep{sankararaman2014genomic, sankararaman2016combined}.

There is growing genetic evidence that early \emph{H. sapiens} reciprocally
contributed to Neanderthal genomes \citep{kuhlwilm2016ancient,
hubisz2020mapping, harris2023diverse}. Such introgression would have occurred
tens or hundreds of thousands of years prior to Neanderthal introgression in
humans during the global dispersal of modern humans around 60 ka. This is
supported by ``near modern'' \emph{H. sapiens} outside of African around
120--100 ka or earlier \citep{schwarcz1988esr, grun2005u, beyer2021climatic},
potentially overlapping with Neanderthals and providing opportunities for early
contacts. While estimates of the genomic contribution of early \emph{H.
sapiens} to Neanderthals vary, around $6\%$ of later Neanderthal genomes may
have been contributed through introgression from humans to Neanderthals
\citep{harris2023diverse}. If \emph{H. sapiens}-related haplotypes carried
fewer deleterious alleles due to their larger long-term effective population
size, human-introgressed DNA would have been favored in Neanderthal genomes.
The replacement of Neanderthal mitochondrial and Y chromosomes by early human
haplotypes supports this model of post-admixture selection in the Neanderthal
lineage \citep{posth2017deeply, petr2020evolutionary}.

Models for selection against introgressed alleles are often based on load
arguments (in particular, differences in the rate of accumulation of
unconditionally deleterious variation in populations of different sizes) or
hybrid incompatibilities \citep{muller1942isolating}. These arguments, founded
in population genetics theory, rarely take into account selection operating on
phenotypic variation or the relationship between genetic and phenotypic
variation. Many phenotypic traits are thought to be under stabilizing selection
\citep{sanjak2018evidence, sella2019thinking}, including gene regulation
\citep{gilad2006natural, hodgins2015gene, price2022detecting}. Because some of
the strongest signals of selection against Neanderthal-introduced alleles are
in regulatory regions \citep{sankararaman2014genomic}, stabilizing selection on
quantitative traits is particularly relevant to the dynamics of functional
genetic variation after introgression.

Stabilizing selection acts to maintain the phenotypic distribution of a trait
near some optimum, which is achieved by reducing phenotypic variation
(Figure~\ref{fig:stab-sel}A). When the mean phenotype of the population is
close to the phenotypic optimum, classical models predict that trait-affecting
alleles are subject to underdominant selection, i.e., selection against the
minor allele at the locus \citep{robertson1956effect}. This has proven to be a
useful model for understanding genetic architectures of traits under
stabilizing selection in single-population settings
\citep[e.g.,][]{keightley1988quantitative, simons2018population,
hayward2022polygenic}.

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/stab_sel.pdf}
    \caption{
        \textbf{Additive genetic variance under stabilizing selection.}
        (A) Stabilizing selection acts to maintain phenotypic values of individuals
        in the population near some optimum. Throughout, we assume a Gaussian fitness
        function.
        (B) With low mutational variance ($V_M$), the expected additive genetic
        variance ($V_G$) is proportional to the population-scaled mutation rate.
        When $V_M$ is large so that mutational effects can be strong, $V_G$ is
        independent of the mutation rate. The stochastic house-of-cards model
        (Eq.~\ref{eq:SHC}) interpolates these regimes, assuming steady-state
        dynamics \citep{burger1989much}. Expected $V_G$ computed using an SFS
        \citep[using \moments,][]{jouganous2017inferring} matches
        simulations assuming no linkage between loci affecting the trait, which
        align closely the stochastic house-of-cards approximation. Here, the
        mutation rate $\mu=0.01$ (per haploid), $V_S=1$, and $N_e=10,000$.
        Mutation effects are normally distributed with mean zero and variance
        $V_M$.
    }
    \label{fig:stab-sel}
\end{figure}

In diverged populations, the genetic variation contributing to a trait under
stabilizing selection has a higher rate of turnover compared to neutrally
evolving loci \citep{yair2022population}. We might therefore expect a rapid
divergence of trait architecture and an accumulation of fixed differences in
the \emph{H. sapiens} and Neanderthal lineages at trait-contributing loci, even
when the mean phenotype in each population remains close to the same trait
optimum. When a derived allele at high frequency in one population is
introduced to another population in which is was previously absent, it will be
at low frequency (if the admixture proportion is low) and subsequently selected
against. Likewise, if the ancestral allele is reintroduced to a population
fixed for a derived allele, the ancestral allele will be at low frequency and
will be selected against. In either case we should expect selection against the
introgressed allele, whether ancestral or derived and regardless the historical
relative sizes of the populations involved. This prediction contrasts with the
population-genetics ``load'' model, in which haplotypes with fewer deleterious
variants (such as those from the population with larger historical size) are
favored after introgression in either direction.

In this article, we show that admixture between diverged populations results in
selection against the minor parental ancestry at trait-affecting loci. We
develop a numerical approach based on the site frequency spectrum to predict
genetic variation under complex demographic scenarios, and which we use to
partition predicted trait heritability by introgressed vs. non-introgressed
variation. Using simulations with linkage, we demonstrate that deserts of
introgressed ancestry form around trait-contributing loci. When gene flow
occurs bidirectionally, such deserts will tend to overlap in location in each
population. As we argue in the Discussion, stabilizing selection on shared
trait optima provides an explanation for the overlap of introgressed ancestry
deserts in human and Neanderthal genomes after reciprocal introgression,
recently observed in \citet{harris2023diverse}.

@@

We consider a polygenic trait for which an individual's additive genetic value
is the sum over all effects of alleles in their genome. For individual $i$,
\(G_i = \sum_l g_{l, i} a_l\), where $a_l$ is the effect size of the derived
allele at locus $l$, and \(g_{l,i}\in\{0,1,2\}\) is their genotype at that
locus (i.e., the number of alleles they carry at that locus). With no linkage
between trait-affecting loci, the expected additive genetic variance is \(V_A =
\sum_l 2p_l(1-p_l)a_l^2\), where $p_l$ is the allele frequency at locus $l$. We
ignore dominance and epistatic effects, so that \(V_G = V_A\). We further
ingore environmental effects \citep{simons2018population}, so that the
phenotypic variance \(V_P=V_G\).

Stabilizing selection acts to reduce phenotypic variation around the optimum
value $O$ (typally set to zero), and we assume a Gaussian fitness function
(Figure~\ref{fig:stab-sel}A) so that relative fitness is given by \(f(G_i | O,
V_S) = \exp{(-(G_i - O)^2 / 2 V_S)}\). $\frac{1}{V_S}$ is interpreted as the
strength of selection on the trait, so larger $V_S$ implies weaker selection.
For a population with mean phenotype at (or very close to) the optimum, the
mean fitness of the population (assuming a roughly normal distribution of
phenotypic values in the population) is
\[\bar{w} \approx \int_{-\infty}^\infty
f(G | 0, V_S) \mathcal{N}(0, V_G) dG =
\left(\frac{V_S}{V_S+V_G}\right)^{1/2}.\]
Thus, as the genetic variance increases, mean fitness among individuals in the
population decreases.

@@

If all alleles contribute equally to the trait
(with effect sizes \(\pm a\) occurring in equal proportion),
\citet{keightley1988quantitative} showed that the dynamics of $V_G$ can be
approximated with the recursion
\[V_{G,t+1} \approx V_{G,t}\left(1-a^2/2(V_S+V_{G,t})\right)
\left(1-1/2N_e\right) + 2 \mu a^2,\]
where $\mu$ is the per-haploid, per-generation rate of mutation. In the
large-population-size limit, this gives the well-known result for
steady-state additive genetic variance,
\[V_G \approx 4\mu V_S,\] provided \(V_G \ll V_S\).

Mutations are not expected to each have the same effect size $|a|$, but rather
follow some distribution. Here, when mutation effects are not constant, we
assume effect sizes are drawn from a normal distribution with mean 0 and given
variance $V_M$. When population sizes or effect sizes are small, drift
dominates the dynamics of $V_G$, and at steady state \citet{lande1976natural}
found \[V_G \approx 4 N_e \mu V_M.\] Interpolating between the selection- and
drift-dominated regimes,
\begin{align}\label{eq:SHC}
    V_G & \approx \frac{4 \mu V_S}{1 + \frac{V_S}{N_e V_M}}.
\end{align}
This, the ``stochastic
house-of-Cards'' (SHC) approximation (Figure~\ref{fig:stab-sel}B), was given by
\citet{burger1989much} and is discussed in detail in \citet[][Ch.
28]{walsh2018evolution}.

@@

As initially shown by \citet{robertson1956effect} \citep[see
also,][]{keightley1988quantitative, simons2018population}, when the mean
phenotype of the population is close to the optimal phenotypic value,
stabilizing selection results in symmetric underdominant selection on alleles
contributing additively to the trait. Thus, for an allele at frequency $p$ with
selection coefficient $s$, the expected change in $p$ over one generation is
\begin{equation}
    \E[\Delta p] = -sp(1-p)(1-2p).
\end{equation}
The selection coefficent $s$ depends on the strength of selection on the trait
$V_S$ and the effect size $a$, as 
\begin{equation}
    s\approx \frac{a^2}{2(V_S+V_G)} \approx \frac{a^2}{2V_S}.
\end{equation}
Details are shown in Appendix~\ref{sec:underdominance}.

@@

We model allelic dynamics using the diffusion approximation, where the expected
change in mean allele frequency per generation is
\[\E[M_{\delta_p}] = E[\Delta p] \approx -s p(1-p)(1-2p),\]
as shown above, and the expected change in variance of the allele frequency per
generation is
\[\E[V_{\delta_p}] \approx \frac{1}{2N_e}p(1-p).\]
Because $s$ is always positive for any $a\not=0$, we see that selection pushes
allele frequencies to zero if $p<1/2$ and to one if $p>1/2$, resulting in
symmetric underdominance.

We extend the \moments-based solution for the sample site-frequency spectrum
(SFS) \citep{jouganous2017inferring} to include underdominance with given
selection coefficient $s$ (Appendix~\ref{sec:moments-underdominance}). The
contribution of alleles with effect size $a$ to the total genetic variance
$V_G$ is found by computing the expected pairwise diversity from the SFS (with
sample size $n$, denoted $\Phi_n$), as \[V_{G,a} \approx 2a^2\sum_{j=1}^{n-1}
\frac{j(n-j)}{n(n-1)} \Phi_n(j|a,\mu_a) da.\] Here, $\mu_a$ is the mutation
rate of alleles with effect size $a$. Then assuming a normal distribution of
effect sizes for new mutations, the total genetic variance is \[V_G \approx
\int_{-\infty}^\infty V_{G,a}\mathcal{N}(0,V_M).\]

If $V_G$ is non-negligible compared to $V_S$, ignoring $V_G$ and using
$s=a^2/2V_S$ leads to deflated estimates of additive genetic variance
(Figure~\ref{fig:supp-one-pop}). When mutation rates are large so that $V_G$ is
not small compared to $V_S$, using $s=a^2/2(V_S+V_G)$ provides estimates of
$V_G$ that closely match simulations assuming free recombination between loci
(Figures~\ref{fig:toy-admixture} and
\ref{fig:one-pop-0.01}--\ref{fig:one-pop-0.1}). Because $V_G$ can vary over
time, this means that $s$ is no longer constant and can change due to factors
such as non-constant demography that increase or reduce $V_G$.

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/one_pop.pdf}
    \caption{
        \textbf{Population size changes and additive genetic variance.}
        We used \moments to track expected genetic variance of a trait
        under stabilizing selection as the population chages size. (A) The
        population goes through a 10-fold size reduction, followed by a recovery.
        (B) All mutations had effect sizes $\pm0.01$, $0.04$, or $0.1$, with equal
        probabilitiy of being trait increasing or decreasing. Depending on the
        selection coefficient ($s\approx a^2/2V_S$) compared to $N_e$, $V_G$
        could increase or decrease after a sudden population size change.
        Predictions using \moments match simulations with free recombination
        between trait-affeting loci
        (Figures~\ref{fig:one-pop-0.01}--\ref{fig:one-pop-0.1}).
    }
    \label{fig:one-pop}
\end{figure}

@@

The second model is meant to more closely resemble inferred human-Neanderthal
history, in which the ancestral population of size $N_e=10{,}000$ splits into the
human and Neanderthal branches. At 250ka, an early human-to-Neanderthal
introgression contributes $5\%$ ancestry to Neanderthals. The human branch
shrinks to size $1{,}000$ 60ka, followed by exponential growth to size $20,000$
at present time. Neanderthals contribute $2\%$ ancestry to this bottlenecked
and expanding human population at 50ka, after which they go extinct
(Figure~\ref{fig:human-neand-h2}A).

In each scenario, we track genetic diversity and phenotypic variance before and
after admixture to understand allele frequency and haplotype dynamics. The
trait optimum is kept at $0$ in all branches, so that no optimum shift occurs,
and the strength of selection $V_S=1$ remains constant.

>>>>

\usepackage[letterpaper,margin=1in]{geometry}
\usepackage{lineno}   %% <- no mathlines option
\usepackage{amsmath}  %% <- after lineno
\usepackage{etoolbox} %% <- for \cspreto, \csappto
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{color}
\usepackage{url}
\usepackage{authblk}
\usepackage{bm}

@@

\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\Cov}{Cov}
\newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
\newcommand{\E}{\mathbb{E}}
\usepackage{xspace}
\newcommand{\moments}{\texttt{moments}\xspace}
\newcommand{\fwdpy}{\texttt{fwdpy11}\xspace}
\newcommand{\tskit}{\texttt{tskit}\xspace}

@@

\title{Archaic introgression and the distribution
of shared variation under stabilizing selection}
\author[]{Aaron P. Ragsdale}
\affil[]{Department of Integrative Biology, University of Wisconsin--Madison, WI, USA}
\date{\today}

@@

    Many phenotypic traits are under stabilizing selection, which maintains a
    population's mean phenotypic value near some optimum. The dynamics of
    traits and trait architectures under stabilizing selection have bene
    extensively studied for single populations at steady state. However,
    natural populations are seldom at steady state. Instead, they are often
    structured in some way, and admixture and introgression events may be
    common, including over human evolutionary history. Because stabilizing
    selection results in selection against the minor allele at a
    trait-affecting locus, alleles from the minor parental ancestry will be
    selected against after admixture. Here, we show that the site-frequency
    spectrum can be used to model the genetic architecture of such traits,
    allowing for the study of trait architecture dynamics in complex
    multi-population settings. We use a simple deterministic two-locus model to
    predict the reduction of introgressed ancestry around trait-contributing
    loci. From this and individual-based simulations, we show that
    introgressed-ancestry deserts are enriched around such loci. When
    introgression between two diverged populations occurs in both directions,
    as has been inferred between humans and Neanderthals, the locations of
    introgressed-ancestry deserts will tend to be shared across populations. We
    argue that stabilizing selection for shared phenotypic optima may explain
    recent observations in which regions of depleted human-introgressed
    ancestry in the Neanderthal genome overlap with Neanderthal-ancestry
    deserts in humans.

@@

Genomic surveys of natural systems show that historical admixture among
diverged populations and closely related taxa commonly occurs
\citep{brandvain2014speciation, skoglund2015ancient, suvorov2022widespread} and
is widespread in primate \citep{tung2017contribution, sorensen2023genome} and
hominin \citep{wolf2018outstanding, peter2020100} evolution. Admixture is
therefore a frequent driver of phenotypic and molecular variation and can
contribute to the genetic architectures of complex traits. In humans, archaic
introgression from Neanderthals and Denisovans has attracted considerable
attention, including efforts to describe the historical processes leading to
observed distributions of introgressed DNA in present-day populations
\citep{prufer2014complete, villanea2019multiple, chen2020identifying} and the
contribution of introgressed variation to quantitative traits
\citep{sankararaman2016combined, wei2023lingering}.

Once introduced through admixture, introgressed alleles may be selected for or
against. Some introgressed haplotypes were likely positively selected in modern
\emph{Homo sapiens} (here, ``humans'') \citep{huerta2014altitude,
racimo2017signatures, enard2018evidence, gower2021detecting}, possibly due to
locally adaptive variation that provided fitness advantages as humans
encountered novel environments. Despite some cases of adaptive introgression,
most introduced alleles were likely to have been selected against in humans
\citep{harris2016genetic, juric2016strength, veller2023recombination}. Since
Neanderthal and Denisovan population sizes were relatively small for hundreds
of thousands of years, theory predicts they would have accumulated deleterious
variation at an increased rate. Introgressed haplotypes loaded with more
deleterious mutations would have been rapidly removed by selection after
admixture. Mapping the distribution of Neanderthal-introgressed haplotypes in
humans shows a reduction of Neanderthal-related ancestry in coding and
regulatory regions \citep{petr2019limits, telis2020selection,
yermakovich2023long}. These ``deserts'' of Neanderthal ancestry support the
hypothesis that introgressed functional alleles were selected against
\citep{sankararaman2014genomic, sankararaman2016combined}.

There is growing genetic evidence that humans reciprocally contributed to
Neanderthal genomes \citep{kuhlwilm2016ancient, hubisz2020mapping,
harris2023diverse, li2024recurrent}. This gene flow occurred tens to hundreds
of thousands of years prior to Neanderthal introgression in humans during the
global dispersal of modern humans around 60 ka. This is supported by early
\emph{H. sapiens} outside of African around 200--100 ka \citep{schwarcz1988esr,
grun2005u, harvati2019apidima, beyer2021climatic}, potentially overlapping with
Neanderthals and providing opportunities for early contacts. While estimates of
the genomic contribution of early humans to Neanderthals vary, around $6\%$ of
later Neanderthal genomes may trace through this admixture event
\citep{harris2023diverse}. Under a ``load'' model, if human-related haplotypes
carried fewer deleterious alleles due to their larger long-term effective
population size, human-introgressed DNA would have been favored in Neanderthal
genomes. The replacement of Neanderthal mitochondrial and Y chromosomes by
early human haplotypes appears to support this model of post-admixture positive
selection in the Neanderthal lineage \citep{posth2017deeply,
petr2020evolutionary}.

Models for selection against introgressed alleles are often based on load
arguments (that is, due to differences in the rate of accumulation of
unconditionally deleterious variation in populations of different sizes) or
hybrid incompatibilities \citep{muller1942isolating}. These models, founded in
population genetics theory, rarely take into account selection operating on
phenotypic traits or the relationship between genetic and phenotypic variation.
Many phenotypic traits are thought to be under stabilizing selection
\citep{sanjak2018evidence, sella2019thinking}, including gene regulation
\citep{gilad2006natural, hodgins2015gene, price2022detecting}. Because some of
the strongest signals of selection against Neanderthal-introduced alleles are
in regulatory regions \citep{sankararaman2014genomic}, stabilizing selection on
quantitative traits may be particularly relevant to the dynamics of functional
genetic variation after introgression among hominins.

Stabilizing selection acts to maintain the phenotypic distribution of a trait
near some optimum, which is achieved by reducing phenotypic variation
(Figure~\ref{fig:stab-sel}A). When the mean phenotype of the population is
close to the phenotypic optimum, classical models predict that the minor allele
at a trait-affecting locus is selected against, with allele-frequency dynamics
equivalent to underdominant selection \citep{robertson1956effect}. This has
proven to be a useful model for understanding genetic architectures of traits
under stabilizing selection in single-population settings
\citep[e.g.,][]{keightley1988quantitative, simons2018population,
hayward2022polygenic}.

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/stab_sel.pdf}
    \caption{
        \textbf{Additive genetic variance under stabilizing selection.}
        (A) Stabilizing selection acts to maintain phenotypic values of
        individuals in the population near some optimum. Throughout, we assume
        a Gaussian fitness function.
        (B) With low mutational variance ($V_M$), the expected additive
        genetic variance ($V_G$) is proportional to the
        population-scaled mutation rate. When $V_M$ is large so that
        mutational effects can be strong, $V_G$ is independent of the mutation
        rate. The stochastic house-of-cards model (Eq.~\ref{eq:SHC})
        interpolates these regimes, assuming steady-state dynamics
        \citep{burger1989much}. Expected $V_G$ computed using an SFS approach
        \citep[developed here using \moments,][]{jouganous2017inferring}
        matches simulations assuming no linkage between loci affecting the
        trait, which align closely with the stochastic house-of-cards
        approximation.
        Here $V_S=1$, $N_e=10{,}000$, the mutation rate
        $\mu=0.01$ (per haploid), and mutation effects are drawn from a normal
        distribution with mean zero and variance $V_M$.
    }
    \label{fig:stab-sel}
\end{figure}

In diverged populations, the genetic variation contributing to a trait under
stabilizing selection has a higher rate of turnover compared to neutral
evolution \citep{yair2022population}. We therefore expect a rapid divergence of
trait architectures and an accumulation of fixed differences in the human and
Neanderthal lineages at trait-contributing loci, even when the mean phenotype
in each population remains close to the same trait optimum. When a derived
allele at high frequency in one population is introduced to another population
in which it was previously absent, it will be at low frequency (if the
admixture proportion is low) and subsequently selected against. Likewise, if
the ancestral allele is reintroduced to a population fixed for a derived
allele, the ancestral allele will be at low frequency and will be selected
against. Selection acts against the introgressed allele in either case, whether
it is ancestral or derived and regardless of the historical relative sizes of
the populations involved. In concurrent work, \citet{veller2024stabilizing}
demonstrate this effect by deriving the expected decay of minor parental
ancestry under stabilizing selection after admixture. This prediction contrasts
with the population-genetics load model, in which haplotypes with fewer
deleterious variants (such as those from the population with larger historical
size) are favored after introgression in either direction.

In this article, we show that admixture between diverged populations results in
selection against the minor parental ancestry at loci contributing to a trait
under stabilizing selection. We develop a numerical approach based on the
site-frequency spectrum to predict genetic variation under complex demographic
scenarios, and which we use to partition predicted trait heritability by
introgressed and non-introgressed variation. Using simulations with linkage, we
demonstrate that deserts of introgressed ancestry form around
trait-contributing loci. When gene flow occurs bidirectionally, such deserts
will tend to overlap in location across populations. We argue that stabilizing
selection on shared trait optima may explain the overlap of introgressed
ancestry deserts in human and Neanderthal genomes after reciprocal
introgression, as has been recently observed \citep{harris2023diverse}.

@@

We consider a polygenic trait for which an individual's additive genetic value
is the sum over all effects of alleles in their genome. For individual $i$,
\(G_i = \sum_l g_{l, i} a_l\), where $a_l$ is the effect size of the derived
allele at locus $l$, and \(g_{l,i}\in\{0,1,2\}\) is their genotype at that
locus (i.e., the number of derived alleles they carry at that locus). With
linkage equilibrium between trait-affecting loci, the expected additive genetic
variance is \(V_A = \sum_l 2p_l(1-p_l)a_l^2\), where $p_l$ is the allele
frequency at locus $l$. We ignore dominance and epistatic effects (often argued
to be a reasonable modeling choice, e.g., \citet{barton2002understanding,
zhang2002pleiotropic}), so that \(V_G = V_A\). We further ignore environmental
effects and noise \citep[under the assumption of additive environmental
effects, as in][]{lande1975maintenance, simons2018population}, so that only
genetic effects are considered.

Stabilizing selection acts to reduce phenotypic variation around the optimum
value $O$, typically set to zero, and we assume a Gaussian fitness function
(Figure~\ref{fig:stab-sel}A) so that relative fitness (relative to an
individual with optimal phenotype) is given by \(f(G_i | O,
V_S) = \exp{(-(G_i - O)^2 / 2 V_S)}\). $\frac{1}{V_S}$ is interpreted as the
strength of selection on the trait, so larger $V_S$ implies weaker selection.
For a population with mean phenotype at or very close to the optimum, the
mean fitness of the population [assuming a normal distribution of phenotypic
values in the population, \citep{turelli1994genetic, urban2013asymmetric}] is
\[\bar{w} \approx \int_{-\infty}^\infty
f(G | O, V_S) \mathcal{N}(O, V_G) dG =
\left(\frac{V_S}{V_S+V_G}\right)^{1/2},\]
so that as the genetic variance increases, population mean fitness decreases.

@@

If all alleles contribute equally to the trait
with effect sizes \(\pm a\) occurring in equal proportion,
\citet{keightley1988quantitative} showed that the dynamics of $V_G$ can be
approximated with the recursion
\[V_{G,t+1} \approx V_{G,t}\left(1-a^2/2(V_S+V_{G,t})\right)
\left(1-1/2N_e\right) + 2 \mu a^2,\]
where $\mu$ is the per-haploid, per-generation rate of mutation. In the
large-population-size limit, this gives the well-known result for
steady-state additive genetic variance,
\[V_G \approx 4\mu V_S,\] provided \(V_G \ll V_S\).

Mutations will not generally all have the same effect size, but rather are
drawn from some distribution. Here, when modeling mutation effects as
non-constant, we assume effect sizes follow a normal distribution with mean 0
and variance $V_M$. When population sizes or effect sizes are small, drift
dominates the dynamics of $V_G$, and at steady state \citet{lande1976natural}
found \[V_G \approx 4 N_e \mu V_M.\] Interpolating between the drift- and
selection-dominated regimes,
\begin{align}\label{eq:SHC}
    V_G & \approx \frac{4 \mu V_S}{1 + \frac{V_S}{N_e V_M}}.
\end{align}
This, the ``stochastic
house-of-Cards'' (SHC) approximation (Figure~\ref{fig:stab-sel}B), was given by
\citet{burger1989much} and is discussed in detail in \citet[][Ch.
28]{walsh2018evolution}.

@@

As initially shown by \citet{robertson1956effect} \citep[see
also,][]{keightley1988quantitative, simons2018population}, when the mean
phenotype of the population is close to the optimal phenotypic value,
stabilizing selection results in selection against the minor alleles at loci
contributing additively to the trait, with dynamics mirroring symmetric
underdominance. In general, for an allele at frequency $p$ with selection
coefficient $s$, the expected change in $p$ over one generation is
\begin{equation}
    \E[\Delta p] = sp(1-p).
\end{equation}

In our case, the selection coefficient $s$ depends on the strength of selection
on the trait $V_S$ and the effect size $a$, as well as the frequency of the
allele, so that
\begin{equation}
    s\approx -\frac{a^2(1-2p)}{2(V_S+V_G)} 
    = \frac{\left(p-\frac{1}{2}\right)a^2}{V_S + V_G}
    \approx \frac{\left(p-\frac{1}{2}\right)a^2}{V_S},
\end{equation}
when $V_G \ll V_S$. This model assumes linkage equilibrium between
trait-contributing loci. We note that LD is expected to be generated between
loci even in the absence of linkage \citep{bulmer1971effect}, and
\citet{negm2024effect} recently showed how to correct for its effect on the
selection dynamics of a focal allele.

Because $a^2/(V_S+V_G)$ is always positive for any $a\not=0$, we see that
selection pushes allele frequencies to zero if $p<1/2$ and to one if $p>1/2$,
resulting in symmetric underdominance. Details are shown in
Appendix~\ref{sec:underdominance}.

@@

We model allelic dynamics using the diffusion approximation, where the expected
change in mean allele frequency per generation is
\[\E[M_{\delta_p}] = E[\Delta p] \approx s p(1-p),\]
with $s$ defined above, and the expected change in variance of the allele frequency per
generation is
\[\E[V_{\delta_p}] \approx \frac{1}{2N_e}p(1-p).\]

We extended the \moments-based solution for the sample site-frequency spectrum
(SFS) \citep{jouganous2017inferring} to include underdominance with given
selection coefficient $s$ (Appendix~\ref{sec:moments-underdominance}). The
contribution of alleles with effect size $a$ to the total genetic variance
$V_G$ is found by computing the expected pairwise diversity from the SFS (with
sample size $n$, denoted $\Phi_n$), as
\[V_{G,a} \approx 2a^2\sum_{j=1}^{n-1} \frac{j(n-j)}{n(n-1)} \Phi_n(j|a,\mu_a) da.\]
Here, $\mu_a$ is the mutation rate of alleles with effect size $a$. Then
assuming a normal distribution of effect sizes for new mutations, the total
genetic variance is
\[V_G \approx \int_{-\infty}^\infty V_{G,a}\mathcal{N}(0,V_M).\]
$V_G$ can be computed periodically in this way to predict trajectories in
non-equilibrium settings (Figure~\ref{fig:one-pop}).

If $V_G$ is non-negligible compared to $V_S$, ignoring $V_G$ and using
$s=(p-1/2)a^2/V_S$ leads to deflated estimates of additive genetic variance
(Figure~\ref{fig:supp-one-pop}). When mutation rates are large so that $V_G$ is
not small compared to $V_S$, using $s=(p-1/2)a^2/(V_S+V_G)$ provides estimates
of $V_G$ that closely match simulations assuming free recombination between
loci (Figures~\ref{fig:toy-admixture} and
\ref{fig:one-pop-0.01}--\ref{fig:one-pop-0.1}). Because $V_G$ can vary over
time, this means that $s$ is no longer constant and can change due to factors
such as non-constant demography that increase or reduce $V_G$.

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/one_pop.pdf}
    \caption{
        \textbf{Population size changes and additive genetic variance.}
        We used \moments to track expected genetic variance of a trait
        under stabilizing selection as the population changes size. (A) The
        population goes through a 10-fold size reduction, followed by a recovery.
        (B) All mutations have effect sizes $\pm0.01$, $0.04$, or $0.1$, with equal
        probability of being trait increasing or decreasing. Depending on the
        selection coefficient ($s\approx (p-1/2)a^2/V_S$) compared to $N_e$, $V_G$
        could increase or decrease after a sudden population size change.
        Predictions using \moments match simulations with free recombination
        between trait-affecting loci
        (Figures~\ref{fig:one-pop-0.01}--\ref{fig:one-pop-0.1}).
    }
    \label{fig:one-pop}
\end{figure}

@@

The second model is meant to more closely resemble inferred human-Neanderthal
history, in which the ancestral population of size $N_e=10{,}000$ splits at 600
kya into the human and Neanderthal branches, with effective sizes $10{,}000$
and $2{,}000$, respectively. At 250 ka, an early human-to-Neanderthal
introgression contributes $5\%$ ancestry to Neanderthals. The human branch
shrinks to size $1{,}000$ 60 ka, followed by exponential growth to size
$20,000$ at present time. Neanderthals contribute $2\%$ ancestry to this
bottlenecked and expanding human population at 50 ka, after which they go
extinct (Figure~\ref{fig:human-neand-h2}A).

In each scenario, we tracked phenotypic variance and genetic variation to
compare simulations to model predictions. The trait optimum was kept at $0$ in
all populations (no optimum shift occurred), and the strength of selection
$V_S=1$ remains constant.
